\documentclass[11pt,a4paper]{article} 

%------------------------------------------------------------------------------
%	REQUIRED PACKAGES AND  CONFIGURATIONS
%------------------------------------------------------------------------------
% Report style
\usepackage{./configuration-files/reportstyle}

% Do not change configuration-files/config.tex file unless you really know what you are doing. 
\input{configuration-files/config}

% Insert here the info that will be displayed into your Title page 
% -> title of your work
\renewcommand{\title}{Numerical Solvers for the Wave Equation}
% -> author name and surname
\renewcommand{\author}{Roberto Di Lauro}
\newcommand{\authorone}{Thomas Fabbris}
\newcommand{\authortwo}{Alessandro Ghiotto}
\newcommand{\authorthree}{Cristian Rubbi}
% -> MSc course
\newcommand{\course}{High Performance Computing Engineering}
% -> author ID
\newcommand{\ID}{10869236}
\newcommand{\IDone}{11174296}
\newcommand{\IDtwo}{11177200}
\newcommand{\IDthree}{11177799}
% -> academic year
\newcommand{\YEAR}{2025-2026}
% -> course name
\newcommand{\COURSENAME}{Numerical Methods for Partial Differential Equations}
% -> abstract
\renewcommand{\abstract}{
This project presents the development, implementation, and performance analysis of a  numerical solver for the second-order wave equation. The solver is built upon the \textbf{deal.II} finite element library and utilizes \textbf{Trilinos} wrappers for distributed linear algebra and parallel computing. The computational domain is discretized using simplicial meshes generated internally, supporting various polynomial degrees $r$ for the finite element space.

\vspace{0.3cm}
\noindent We investigate two main families of time-integration schemes: the \textbf{$\theta$-methods} (e.g., Forward Euler, Crank-Nicolson, and Backward Euler) and the \textbf{Newmark-beta} family. A rigorous numerical analysis is performed, focusing on:

\vspace{0.1cm}
\begin{enumerate}
    \item \textbf{Convergence Analysis}: Evaluation of relative $L^2$ and $H^1$ errors across multiple spatial and temporal resolutions to verify theoretical accuracy rates.
    \item \textbf{Dissipation and Dispersion}: Study of energy conservation properties and phase errors introduced by the discrete approximation.
    \item \textbf{Scalability}: Performance benchmarks conducted on the \textbf{Politecnico di Milano HPC "Students Cluster"}, evaluating parallel speedup and efficiency.
\end{enumerate}

\vspace{0.3cm}
\noindent The results provide a comprehensive characterization of the solver's reliability and efficiency, demonstrating its suitability for large-scale dynamic wave propagation simulations.
}


%-------------------------------------------------------------------------
%	BEGIN OF YOUR DOCUMENT
%-------------------------------------------------------------------------
\begin{document}

%-----------------------------------------------------------------------------
% TITLE PAGE
%-----------------------------------------------------------------------------
\input{configuration-files/title-page}

%-----------------------------------------------------------------------------
% TABLE OF CONTENTS
\newpage
\tableofcontents
%-----------------------------------------------------------------------------
% INTRODUCTION
\newpage
%-----------------------------------------------------------------------------
\input{sections/1_introduction.tex}

\vspace{1.0cm}

\input{sections/2_Continuous_Wave_Problem.tex}
\vspace{0.5cm}
\input{sections/3_Spatial_Discretization.tex}
\vspace{1.0cm}
\input{sections/4_Time_Discretization.tex}
% \newpage
\input{sections/5_Implementation_Details.tex}
\input{sections/6_Convergence.tex}
\input{sections/7_Dissipation_Dipsersion.tex}
\input{sections/8_Scalability.tex}

%-----------------------------------------------------------------------------
% CONCLUSION
%-----------------------------------------------------------------------------
\input{sections/9_conclusion.tex}
\newpage
%-----------------------------------------------------------------------------
% BIBLIOGRAPHY
%-----------------------------------------------------------------------------
\bibliography{bibliography.bib}

%-----------------------------------------------------------------------------
% APPENDICES
%-----------------------------------------------------------------------------
% \appendix
% \section{Appendix A}
% \section{Appendix B}

%-------------------------------------------------------------------------
%	END OF YOUR DOCUMENT
%-------------------------------------------------------------------------
\end{document}
